A comparative analysis of penalized linear mixed models in structured genetic data

### 2020-12-11 main.tex
Updates from Overleaf
Edits to 1, 3, 4, 5, bibliography, eps figure format

\pb{work this citation in; maybe others like it: https://journals.plos.org/plosgenetics/article?id=10.1371/journal.pgen.1009141}

@@

% \anna{keep this for later use when talking about genetic background?} Genetic variants are often correlated with each other such that univariate testing may be prone to false positives and biased estimates. When a genetic variant of interest is highly correlated with others, the effects of these will be attributed to the variant in question. This may result in causal SNPs being removed from analysis, so that only SNPs in LD with the causal loci, as opposed to the causal loci themselves, can be identified and their effects estimated.

One method for analyzing data in the presence of correlated loci is LD pruning, which involves sequentially thinning SNPs, so that only those correlated with each other below some desired threshold are used in modeling and testing procedures. LD pruning can be an important preprocessing step in a GWAS analysis, for which there exist a variety of algorithms and methods. An alternative solution to the problem of LD is the use of multivariate methods such as multiple linear regression, which adjusts for and estimates the effects of numerous variants simultaneously. However, classical approaches rely on collecting data with more observations than unknown parameters. This is typically infeasible in genome-wide studies where the number of features greatly exceeds the number of observations. Penalized regression methods make such problems tractable by constraining the solution space of coefficient estimates. The lasso is a widely used penalized regression method \citep{tibshirani1996regression}. It relies on an assumption of sparsity, which allows coefficient estimation and variable selection to be accomplished simultaneously. It also provides considerable computational advantages that scale up well to very large data sets. These features make the lasso attractive for the analysis of high-dimensional genetic data where identifying a list of variants most highly associated with a particular trait is often of interest.

As with multiple linear regression models, lasso penalized regression assumes independent observations and uncorrelated errors. In the presence of population structure or cryptic relatedness, however, samples are correlated, violating this assumption. If unaccounted for, this dependency can result in the identification of spurious associations. Though methods to correct for population stratification and cryptic relatedness have been extensively investigated, until recently the majority of this work has focused on plant and animal breeding data, much of which is experimentally controlled \citep{Amin2007, hoffman2013correcting, price2006principal, Rakitsch2012, bhatnagar2019simultaneous, Sillanpaeae2011}. While many standard correction techniques may in principal be applied to human genetic data, their performance has not been well-studied within the context of human ancestry and non-experimental conditions \citep{lawson2019population, barton2019population}.

@@

In order to clarify and emphasize the importance of the subtle distinction between population structure and environmental heterogeneity, we present a detailed review of relevant concepts and methods. The remainder of this paper is organized as follows. In Section \ref{sec:background} we provide a summary of the causes and consequences of structure in the genetic data of seemingly unrelated individuals and formally define terminology used throughout this paper. We emphasize that confounding due to population structure, frequently cited as a driver of spurious associations, is a function of environmental heterogeneity, where genetic data serves as a proxy for differential environmental exposures \citep{Sillanpaeae2011, sul2018population, vilhjalmsson2012nature, barton2019population}. In so doing, we refine the definition of confounding due to population structure in terms of a non-genetic mechanism. In addition, we briefly review historical methods of correcting for population stratification and relatedness. In Section \ref{sec:methods} we provide a detailed review of the two most prevalent multivariate methods of adjusting for population stratification and relatedness at present: PCA adjusted lasso penalized regression (PC-lasso) and lasso penalized multivariate LMMs (LMM-lasso). We review the statistical details of how these methods attempt to correct for population structure and environmental heterogeneity. In Section \ref{sec:results}, we illustrate the concepts described in Section \ref{sec:methods} via simulation studies, where the data-generating mechanism formally differentiates between population structure and environmental heterogeneity. In addition, these results identify scenarios where particular methods may be most and least effective in accurately estimating the effect sizes of observed SNPs in the presence of unobserved confounding.

@@

Population structure is defined by the existence of allele frequency differences that define sub-populations and is driven by the combined effects of evolutionary processes such as genetic drift, migration, demographic history, and natural selection \citep{gibson2015primer, tibayrenc2017genetics}. Population structure is broadly categorized based on whether it describes recent or ancient relatedness. Ancient relatedness describes the presence of a common ancestor many generations previously. The presence of distinct ancestry groups with different allele frequencies in a sample is known as \textbf{population stratification}. Recent relatedness describes the sharing of a common ancestor only several generations previously. Pedigree-based methods may be used to explicitly model recent relatedness if familial relationships are known. In the absence of known familial relationships, recent relatedness is referred to as \textbf{cryptic relatedness.}

@@

Whether or not such methods are effective at reducing confounding effects depends upon whether population structure can be accurately inferred. An important concept to note, and one that we will return to in Section \ref{sec:partition}, is that population structure only leads to confounding if it is correlated with the environment.
In the presence of environmental heterogeneity which affects the phenotype of interest in a manner that is orthogonal to population structure, SNP-phenotype associations will not be confounded. This is due to the absence of a causal pathway between population stratification and environment, as depicted in diagram \eqref{diag:ps_env_block2}. In such a scenario, failing to account for unobserved environmental effects is likely to result in greater variability, but SNP effect estimates will remain unbiased \citep{greenland1999causal}. We will return to this idea and its mathematical justification in Section \ref{sec:methods}.

@@

Assuming for now that population structure is correlated with environmental exposures and can be accurately inferred from available data, we return to our consideration of methods that attempt to control for confounding via the mechanism depicted in diagram \eqref{diag:ps_env_block}. Such methods include LMMs, principal-component-based approaches, and stratified analyses. Stratified analyses refer to a general class of methods which attempt to infer subject membership within a discrete number of nonoverlapping subpopulations. Subsequent analyses are conducted within each subpopulation, and the subpopulation-specific results may then be combined via meta-analysis \citep{pritchard1999use, pritchard2000association}. By partitioning the overall sample size in this manner, the structured association method is vulnerable to a loss of power. Additionally, it relies on the assumption that these subpopulations are distinct and nonoveralapping. This assumption may not accurately reflect the characteristics of a particular sample, such as one consisting of admixed populations, or recent relatedness. Considered in the context of our air quality example, rather than the existence of two subpopulations with exposure to either good or bad air quality, there may also be individuals who fall between these two extremes and are characterized by continuously varying relatedness and a gradient of air quality. By clustering those individuals within in one subpopulation or another, we lose information about relatedness by oversimplifying the population structure.

@@

% rely on a pair-wise similarity matrix
Like PC-adjustment methods, LMMs utilize the left singular vectors of a similarity matrix to estimate and subsequently correct for relatedness among subjects. In an LMM, however, these left singular vectors are treated as random, rather than fixed effects. The term LMM has been used to describe a number of distinct but related procedures for analyzing structured genetic data which differ in their objectives and methods. There are two main objectives of LMMs used in a structured genetic context: estimating narrow-sense heritability, and estimating individual SNP effects.

@@

Multivariate mixed model approaches reduce these problems by modeling the relationship between a phenotype of interest and all SNPs simultaneously via penalized regression, while correcting for dependencies among subjects \citep{Rakitsch2012, bhatnagar2019simultaneous}. This approach assumes there are a relatively small number of SNPs associated with the phenotype of interest that have large effect sizes. These SNPs will be included in the model as both fixed and random effects, while SNPs with smaller effect sizes are modeled only as part of the random effect in order to control for dependent errors due to relatedness and sample structure.

The remainder of this review will focus on the performance of such multivariate lasso-penalized methods for the analysis of continuous, normally distributed outcomes. Specifically, PC-adjustment and LMM methods, which we will refer to as PC-lasso and LMM-lasso, respectively. We consider two similar implementations of the LMM-lasso method, one developed by \citet{Rakitsch2012} and a more recent implementation from \citet{bhatnagar2019simultaneous}. Subsequent references to LMM-lasso apply to both implementations, unless otherwise stated. Where necessary, we will differentiate between these implementations by referring to them as LMM-lasso-Rakitsch and LMM-lasso-ggmix, respectively.

@@

Both PC-lasso and LMM-lasso rely on matrix decompositions of $\bX$ and the realized relationship matrix (RRM), $\bK = p^{-1} \bX \bX^\T$, which describes the pair-wise genetic similarity among subjects \citep{patterson2006population, hayes2009increased}. Using the singular value decomposition (SVD) of the standardized genotype matrix, we can derive the spectral decomposition of the RRM:

\begin{equation}
    \label{eqn:k}
    \bK = \frac{1}{p} \bX \bX^\T = \frac{1}{p} \bU \bD^2 \bU^\T = \frac{1}{p} \bR \bR^\T,
\end{equation}
where $\bD^2$ is an $n \times n$ diagonal matrix containing the $k$ nonzero eigenvalues of $\bK$ on the top-left diagonal, followed by $n - k$ zeroes on the bottom-right diagonal. $\bU$ is the corresponding $n \times n$ orthonormal matrix of left eigenvectors, or principal components, of $\bK$. The first $k$ columns of $\bU$ correspond to the nonzero eigenvalues of $\bK$, and the columns of $\bR$ are thus the principal components of $\bK$ weighted by their corresponding singular values.

@@

Different variance component estimation methods lead to different forms of $\widehat{\bW}$ and two distinct LMM-lasso variants: LMM-lasso-Rakitsch and LMM-lasso-ggmix. LMM-lasso-Rakitsch uses a two-step procedure for estimating the variance components, $\sigmaee$ and $\delta := \sigmaee / \sigmagg$ \citep{Rakitsch2012}. These variance components and the corresponding weights are estimated once under the assumption of a null model. A lasso model can then be fit on the transformed data to estimate $\boldsymbol{\beta}$ using standard software for coordinate descent, such as \texttt{glmnet} \citep{glmnet} or \texttt{ncvreg} \citep{ncvreg}. LMM-lasso-ggmix iteratively updates $\widehat{\boldsymbol{\beta}}$ and the variance components via a block coordinate descent algorithm \citep{bhatnagar2019simultaneous}. This iterative update scheme prevents the use of standard lasso software, but is implemented in the \texttt{R} package \texttt{ggmix} \citep{ggmix}. We further discuss differences between these methods in Section \ref{sec:results}.

%------------------------------%
% \subsubsection{\anna{Intuition from GLS (?) p-values and SEs in response to rotation?}}

@@

%We argue that this is somewhat misleading from the perspective of understanding the statistical properties of the method in the presence of environmental heterogeneity. Averaging over situations in which the environmental confounding is constantly changing direction and magnitude has the effect of ``washing out'' bias in $\bbetaHat$.

@@

Finally, it is worth notiong that the proportion of environmental effects in the column space of $\P_{\bX}$ (i.e., the portion that will cause confounding) is given by
\begin{equation}
    \label{eqn:ratio}
    \text{CP} = ||\bX \boldsymbol{\tau}||_2^2 / ||\bZ \bgamma||_2^2,
\end{equation}
which we refer to as the Confounding Proportion (CP). The CP can range from 0 (no confounding) to 1. In the extreme case where $\bZ \bgamma$ is completely independent of population structure, $\bZ \bgamma \in \mathcal{N}(\P_{\bX})$ and $\bX \boldsymbol{\tau} = \boldsymbol{0}$ so that $\mathbb{E}(\bbetaHat^*) = \bbeta$. In this scenario, the SNP-phenotype associations will be unbiased, although the residual error will be inflated.

@@

We generated these synthetic phenotypes using three types of high-dimensional ($p > n$) $\bG$ matrices, which were then standardized to yield the corresponding $\bX$ matrix: (1) simulated SNP data from admixed subpopulations, (2) simulated SNP data from independent subpopulations, and (3) real genome-wide SNP data from known caucasian ($n = 417$), African American ($n = 134$), and Hispanic ($n = 37$) subpopulations \citep{larkin2015objectives}. We refer to these as the 1D Linear Admixture, Independent Subpopulations, and Empirical data, respectively. For all data types $p = 1000$ with 50 causal SNPs. For the simulated 1D Linear Admixture and Independent Subpopulations data $n = \{200, 400, 600, 800 \}$, and $n = 588$ for the Empirical data.

@@

In our simulations, $\bZ$ is a matrix of indicator functions identifying subpopulation (founding population) membership, which assigns the corresponding environmental effect $\bgamma$. Various $\bgamma$ patterns were considered in order to investigate the effects of distinct patterns of environmental confounding.  Simulations were carried out with dichotomous (high/low) environmental effects, as well as more complex heterogeneous effects of all different sizes.  In addition, we examined the ``concordance'' of the environmental effect: whether individuals in genetically similar subpopulations experience similar environmental effects or not.  In our experiments, however, neither the heterogeneity nor the concordance of the environmental effects had any meaningful impact on the performance of the methods under consideration.  Thus, throughout Section~\ref{sec:results}, we present results only for disconcordant, dichotomous environmental effects, as the other results are very similar.

%Firstly, we considered two different patterns of effect magnitudes, which we refer to as exponential and dichotomous. We use the term dichotomous to describe two effect sizes each experienced by half of the population, regardless of the number of subpopulations, while exponential effects refer to a unique effect size for each subpopulation with exponentially increasing magnitudes. Secondly, we considered whether individuals in genetically similar subpopulations would experience similar environmental effects. We refer to environmental effects that are similar among genetically similar subpopulations as concordant, and to those that are not as discordant. \anna{We examined all combinations of these patterns of environmental effects, but found they did not result in appreciable differences in performance. We thus focus our attention elsewhere.}

@@

We begin by comparing the ordinary lasso, PC-lasso, and LMM-lasso as the level of environmental confounding is varied. Estimation accuracy for lasso, PC-lasso, and LMM-lasso was evaluated based on mean squared error (MSE) \pb{include a formula}. Figure \ref{fig:mse} presents these results for all three data types in a 1:1 SNR setting, where $\eta = 0.5$. The top row of Figure \ref{fig:mse} shows that when $\xi = 0$ (i.e., no environmental effects), lasso and LMM-lasso perform comparably. PC-lasso results in a slight inflation in MSE compared to the other methods in the simulated data sets due to the fact that 10 additional predictors (PCs) have been added to the model, but none of them actually explain any variability in the outcome.

\begin{figure}[H]
    \centering
    \includegraphics[width = \textwidth]{figures/beta_mse.png}
    \caption{Estimation accuracy of different methods for different data types in the presence and absence of environmental confounding with coarse subpopulation structure and dichotomous-discordant environmental effects.}
    \label{fig:mse}
\end{figure}

@@

In order to assess the behavior of these methods in this context, we conducted a simulation based on the Empirical genotype type, but restricted to just two subpopulations, Caucasian and African-American.  \pb{DESCRIBE HOW SNPS CHOSEN HERE}
As in the earlier simulations, the SNR was set to 1 ($\eta = 0.5$).

 \begin{figure}[H]
    \centering
    \includegraphics[width = \textwidth]{figures/pop_inf_snps.png}
    \caption{\anna{Association of population informative nature of SNPs and false selection across methods in the presence and absence of environmental confounding for two Empirical subpopulations, dichotomous-discordant effects, $\eta = 0.5$.}}
    \label{fig:pop_inf}
\end{figure}

%\begin{figure}[H]
%\begin{subfigure}{\textwidth}
%  \centering
%  % include first image
%  \includegraphics[scale = 0.9]{figures/pop_inf_snps_loess.png}  
%  \caption{FPR in the presence of varying levels of population informative SNPs.}
%  \label{fig:pop_inf_scatter}
%\end{subfigure}
%\begin{subfigure}{\textwidth}
%  \centering
%  % include second image
%  \includegraphics[scale = 0.9]{figures/pop_inf_snps_box.png}  
%  \caption{Absolute allele frequency differences of falsely selected, non-causal SNPs across methods.}
%  \label{fig:pop_inf_box}
%\end{subfigure}
%\caption{Association of population informative nature of SNPs and false selection across methods in the presence and absence of environmental confounding for for two Empirical subpopulations, dichotomous-discordant effects, $\eta = 0.5$.}
%\label{fig:pop_inf}
%\end{figure}

%% \anna{maybe no table ? If table, make more like table 1}
%% \begin{table}[H]
%% \centering
%% \begin{tabular}{rrlr}
%%   \hline
%% eta & xi & Method & $r_s$ \\ 
%%   \hline
%% 0.5 & 0.0 & lasso & 0.42 \\ 
%%   0.5 & 0.0 & PC-lasso & -0.77 \\ 
%%   0.5 & 0.0 & LMM-lasso & -0.71 \\ 
%%   0.5 & 0.8 & lasso & 0.67 \\ 
%%   0.5 & 0.8 & PC-lasso & -0.61 \\ 
%%   0.5 & 0.8 & LMM-lasso & -0.34 \\ 
%%    \hline
%% \end{tabular}
%% \caption{\anna{Spearman's rank correlation coefficient between SNP FPR and absolute difference in standardized allele frequencies across methods.}}
%% \label{tab:pop_inf}
%% \end{table}

%% \anna{Table \ref{tab:pop_inf} displays the Spearman's rank correlation coefficient for each of these scenarios.} In the absence of environmental confounding, the lasso demonstrates a moderately positive correlation with allele frequency differences ($r_s = 0.42$), while the PC-lasso and LMM-lasso methods are negatively correlated with allele frequency differences. Overall, however, the FPR does not exceed 0.39, 0.19, and 0.22 for the lasso, PC-lasso, and LMM-lasso methods, respectively.

@@

%Nearly all of its false positive selections are lasso's correlation with allele frequency increases ($r_s = 0.67$), and particularly struggles to correctly select the most population-informative SNPs. PC-lasso and LMM-lasso are able to remain more robust to allele frequency differences, yet remain somewhat negatively correlated with allele frequency differences. This is advantageous in preventing dramatically high FPRs in the presence of highly population-informative SNPs. From a marginal perspective, LMM-lasso seems to exhibit even stronger control of this than PC-lasso. This echoes the superior TPRs of LMM-lasso compared to those of the PC-lasso for the Empirical data in Table \ref{tab:var_sel}. Conditional on selection, however, LMM-lasso may be more likely to confuse MAF differences with true signal.

@@

\pb{Need to explicitly say how many subpopulations in each case, both in text and legend.}

Although PC-lasso and LMM-lasso outperform the ordinary lasso in the presence of environmental confounding, they are not equally effective across all population structures.  Here, we compare the performance of these two methods in two settings, as described in Section~\ref{sec:subpopulation_structure}: ``coarse'' (\pb{XXX subpopulations}) and ``fine'' (\pb{XXX subpopulations}).  Synthetic phenotypes were simulated using a 1:1 SNR ($\eta = 0.5$) and varying levels of environmental confounding, $\xi \in \{0.2, 0.5,0.8\}$.

\begin{figure}[H]
    \centering
    \includegraphics[scale = 0.9]{figures/mse_diff_subpops.png}
    \caption{Relative performance of LMM-lasso, PC-lasso for fine and coarse subpopulation structures and varying levels of environmental confounding with Independent Subpopulations data, $\eta = 0.5$, and dichotomous-discordant environmental effects. }
    \label{fig:big_vs_small}
\end{figure}

@@

The underlying reason for this is that the RRM becomes ill-conditioned when $n > p$ \citep{ledoit2004well}. Although both approaches utilize the RRM, because LMM-lasso-ggmix attempts to estimate more parameters than LMM-lasso-Rakitsch, its optimization problem is no longer convex, making it prone to becoming stuck in local minima before it is able to sufficiently explore the coefficient path. This phenomenon is illustrated in greater detail in Supplementary Figure \ref{fig:niter}.
%Converging to a local minimum in this manner would lead to few iterations required for each $\lambda$ value before convergence, allowing for an advantage in terms of computational speed, but poor estimation, which is what we observe in Figure \ref{fig:eta_beta_mse}.

\begin{figure}[H]
    \centering
    \includegraphics[scale = 1]{figures/eta_beta_hat.png}
     \caption{$\bbeta$ and $\eta$ estimation accuracy of LMM-lasso-Rakitsch and LMM-lasso-ggmix for varying $\eta$ levels and sample sizes with Independent Subpopulations data, coarse subpopulation structure, $\xi = 0.8$, and dichotomous-discordant environmental effect structure. \pb{Can we make this figure less tall?}}
    \label{fig:eta_beta_mse}
\end{figure}

>>>>

@@

One method for analyzing data in the presence of correlated loci is LD pruning, which involves sequentially thinning SNPs, so that only those correlated with each other below some desired threshold are used in modeling and testing procedures. LD pruning can be an important preprocessing step in a GWAS analysis, for which there exist a variety of algorithms and methods. An alternative solution to the problem of LD is the use of multivariate methods such as multiple linear regression, which adjusts for and estimates the effects of numerous variants simultaneously. However, classical approaches rely on collecting data with more observations than unknown parameters. This is typically infeasible in genome-wide studies where the number of features greatly exceeds the number of observations. Penalized regression methods make such problems tractable by constraining the solution space of coefficient estimates. The lasso is a widely used penalized regression method \citep{tibshirani1996regression}. It relies on an assumption of sparsity, which allows coefficient estimation and variable selection to be accomplished simultaneously. It also provides considerable computational advantages that scale up well to very large data sets. These features make the lasso attractive for the analysis of high-dimensional genetic data where identifying a list of variants most highly associated with a particular trait is often of interest. The lasso is widely used in this context and its application to ultra high-dimensional genetic data is the subject of recent research \citep{qian2020fast, prive2018efficient}.
%The lasso has been increasingly used in this context; of particular note is its use in the analysis of the UK Biobank data \citep{qian2020fast, prive2018efficient}.

As with multiple linear regression models, lasso penalized regression assumes independent observations and uncorrelated errors. In the presence of population structure or cryptic relatedness, however, samples are correlated, violating this assumption. If unaccounted for, this dependency can result in the identification of spurious associations. Though methods to correct for population stratification and cryptic relatedness have been extensively investigated, until recently the majority of this work has focused on plant and animal breeding data, much of which is experimentally controlled \citep{Amin2007, hoffman2013correcting, price2006principal, Rakitsch2012, bhatnagar2020simultaneous, Sillanpaeae2011}. While many standard correction techniques may in principal be applied to human genetic data, their performance has not been well-studied within the context of human ancestry and non-experimental conditions \citep{lawson2019population, barton2019population}.

@@

In order to clarify and emphasize the importance of the subtle distinction between population structure and environmental heterogeneity, we present a detailed review of relevant concepts and methods. The remainder of this paper is organized as follows. In Section \ref{sec:background} we provide a summary of the causes and consequences of structure in the genetic data of seemingly unrelated individuals and formally define terminology used throughout this work. We emphasize that confounding due to population structure, frequently cited as a driver of spurious associations, is a function of environmental heterogeneity, where genetic data serves as a proxy for differential environmental exposures \citep{Sillanpaeae2011, sul2018population, vilhjalmsson2012nature, barton2019population}. In so doing, we refine the definition of confounding due to population structure in terms of a non-genetic mechanism. In addition, we briefly review historical methods of correcting for population stratification and relatedness. In Section \ref{sec:methods} we provide a detailed review of the two most prevalent multivariate methods of adjusting for population stratification and relatedness at present: PCA adjusted lasso penalized regression (PC-lasso) and lasso penalized multivariate LMMs (LMM-lasso). We review the statistical details of how these methods attempt to correct for population structure and environmental heterogeneity. In Section \ref{sec:results}, we illustrate the concepts described in Section \ref{sec:methods} via simulation studies, where the data-generating mechanism formally differentiates between population structure and environmental heterogeneity. In addition, these results identify scenarios where particular methods may be most and least effective in accurately estimating the effect sizes of observed SNPs in the presence of unobserved confounding.

@@

Population structure is defined by the existence of allele frequency differences that define subpopulations and is driven by the combined effects of evolutionary processes such as genetic drift, migration, demographic history, and natural selection \citep{gibson2015primer, tibayrenc2017genetics}. Population structure is broadly categorized based on whether it describes recent or ancient relatedness. Ancient relatedness describes the presence of a common ancestor many generations previously. The presence of distinct ancestry groups with different allele frequencies in a sample is known as \textbf{population stratification}. Recent relatedness describes the sharing of a common ancestor only several generations previously. Pedigree-based methods may be used to explicitly model recent relatedness if familial relationships are known. In the absence of known familial relationships, recent relatedness is referred to as \textbf{cryptic relatedness.}

@@

Whether or not such methods are effective at reducing confounding effects depends upon whether population structure can be accurately inferred. An important concept to note, and one that we will return to in Section \ref{sec:partition}, is that population structure only leads to confounding if it is correlated with distinct environmental effects.
In the presence of environmental heterogeneity which affects the phenotype of interest in a manner that is orthogonal to population structure, SNP-phenotype associations will not be confounded. This is due to the absence of a causal pathway between population stratification and environment, as depicted in diagram \eqref{diag:ps_env_block2}. In such a scenario, failing to account for unobserved environmental effects is likely to result in greater variability, but SNP effect estimates will remain unbiased \citep{greenland1999causal}. We will return to this idea and its mathematical justification in Section \ref{sec:methods}.

@@

Assuming for now that population structure is correlated with environmental exposures and can be accurately inferred from available data, we return to our consideration of methods that attempt to control for confounding via the mechanism depicted in diagram \eqref{diag:ps_env_block}. Such methods include LMMs, principal-component-based approaches, and structured association. Structured association refers to a general class of methods which attempt to infer subject membership within a discrete number of nonoverlapping subpopulations. Subsequent analyses are conducted within each subpopulation, and the subpopulation-specific results may then be combined via meta-analysis \citep{pritchard1999use, pritchard2000association}. By partitioning the overall sample size in this manner, the structured association method is vulnerable to a loss of power. Additionally, it relies on the assumption that these subpopulations are distinct and nonoveralapping. This assumption may not accurately reflect the characteristics of a particular sample, such as one consisting of admixed populations, or recent relatedness. Considered in the context of our air quality example, rather than the existence of two subpopulations with exposure to either good or bad air quality, there may also be individuals who fall between these two extremes and are characterized by continuously varying relatedness and a gradient of air quality. By clustering those individuals within in one subpopulation or another, we lose information about relatedness by oversimplifying the population structure.

@@

Like PC-adjustment methods, LMMs utilize the left singular vectors of a similarity matrix to estimate and subsequently correct for relatedness among subjects. In an LMM, however, these left singular vectors are treated as random, rather than fixed effects. The term LMM has been used to describe a number of distinct but related procedures for analyzing structured genetic data which differ in their objectives and methods. There are two main objectives of LMMs used in a structured genetic context: estimating narrow-sense heritability, and estimating individual SNP effects.

@@

Multivariate mixed model approaches reduce these problems by modeling the relationship between a phenotype of interest and all SNPs simultaneously via penalized regression, while correcting for dependencies among subjects \citep{Rakitsch2012, bhatnagar2020simultaneous}. This approach assumes there are a relatively small number of SNPs associated with the phenotype of interest that have large effect sizes. These SNPs will be included in the model as both fixed and random effects, while SNPs with smaller effect sizes are modeled only as part of the random effect in order to control for dependent errors due to relatedness and sample structure.

The remainder of this review will focus on the performance of such multivariate lasso-penalized methods for the analysis of continuous, normally distributed outcomes. Specifically, PC-adjustment and LMM methods, which we will refer to as PC-lasso and LMM-lasso, respectively. We consider two similar implementations of the LMM-lasso method, one developed by \citet{Rakitsch2012} and a more recent implementation from \citet{bhatnagar2020simultaneous}. Subsequent references to LMM-lasso apply to both implementations, unless otherwise stated. Where necessary, we will differentiate between these implementations by referring to them as LMM-lasso-Rakitsch and LMM-lasso-ggmix, respectively.

@@

Both PC-lasso and LMM-lasso rely on matrix decompositions of $\bX$ and the realized relationship matrix (RRM), $\bK = p^{-1} \bX \bX^\T$, which describes the pair-wise genetic similarity among subjects \citep{patterson2006population, hayes2009increased}. Using the singular value decomposition (SVD) of the standardized genotype matrix, we can derive the spectral decomposition of the RRM: 
\begin{equation}
    \label{eqn:k}
    \bK = \frac{1}{p} \bX \bX^\T = \frac{1}{p} \bU \bD^2 \bU^\T = \frac{1}{p} \bR \bR^\T,
\end{equation}
where $\bD^2$ is an $n \times n$ diagonal matrix containing the $k$ nonzero eigenvalues of $\bK$ on the top-left diagonal, followed by $n - k$ zeroes on the bottom-right diagonal. $\bU$ is the corresponding $n \times n$ orthonormal matrix of left eigenvectors, or principal components, of $\bK$. The first $k$ columns of $\bU$ correspond to the nonzero eigenvalues of $\bK$, and the columns of $\bR$ are thus the principal components of $\bK$ weighted by their corresponding singular values.

@@

Different variance component estimation methods lead to different forms of $\widehat{\bW}$ and two distinct LMM-lasso variants: LMM-lasso-Rakitsch and LMM-lasso-ggmix. LMM-lasso-Rakitsch uses a two-step procedure for estimating the variance components, $\sigmaee$ and $\delta := \sigmaee / \sigmagg$ \citep{Rakitsch2012}. These variance components and the corresponding weights are estimated once under the assumption of a null model. A lasso model can then be fit on the transformed data to estimate $\boldsymbol{\beta}$ using standard software for coordinate descent, such as \texttt{glmnet} \citep{glmnet} or \texttt{ncvreg} \citep{ncvreg}. LMM-lasso-ggmix iteratively updates $\widehat{\boldsymbol{\beta}}$ and the variance components via a block coordinate descent algorithm \citep{bhatnagar2020simultaneous}. This iterative update scheme prevents the use of standard lasso software, but is implemented in the \texttt{R} package \texttt{ggmix} \citep{ggmix}. We further discuss differences between these methods in Section \ref{sec:results}.

@@

@@

Finally, it is worth noting that the proportion of environmental effects in the column space of $\P_{\bX}$ (i.e., the portion that will cause confounding) is given by
\begin{equation}
    \label{eqn:ratio}
    \text{CP} = ||\bX \boldsymbol{\tau}||_2^2 / ||\bZ \bgamma||_2^2,
\end{equation}
which we refer to as the Confounding Proportion (CP). The CP can range from 0 (no confounding) to 1. In the extreme case where $\bZ \bgamma$ is completely independent of population structure, $\bZ \bgamma \in \mathcal{N}(\P_{\bX})$ and $\bX \boldsymbol{\tau} = \boldsymbol{0}$ so that $\mathbb{E}(\bbetaHat^*) = \bbeta$. In this scenario, the SNP-phenotype associations will be unbiased, although the residual error will be inflated.

@@

We generated these synthetic phenotypes using three types of high-dimensional ($p > n$) $\bG$ matrices, which were then standardized to yield the corresponding $\bX$ matrix: (1) simulated SNP data from admixed subpopulations, (2) simulated SNP data from independent subpopulations, and (3) real genome-wide SNP data from known caucasian ($n = 417$), African American ($n = 134$), and Hispanic ($n = 37$) subpopulations \citep{larkin2015objectives}. We refer to these as the 1D Linear Admixture, Independent Subpopulations, and Empirical data, respectively. For all data types $p = 1000$ with 50 causal SNPs. For the simulated 1D Linear Admixture and Independent Subpopulations data $n = 200$ unless otherwise specified, and $n = 588$ for the Empirical data.

@@

In our simulations, $\bZ$ is a matrix of indicator functions identifying subpopulation (founding population) membership, which assigns the corresponding environmental effect $\bgamma$. Various $\bgamma$ patterns were considered in order to investigate the effects of distinct patterns of environmental confounding.  Simulations were carried out with dichotomous (high/low) environmental effects, as well as more complex heterogeneous effects of all different sizes.  In addition, we examined the ``concordance'' of the environmental effect: whether individuals in genetically similar subpopulations experience similar environmental effects or not.  In our experiments, however, neither the heterogeneity nor the concordance of the environmental effects had any meaningful impact on the performance of the methods under consideration.  Thus, throughout Section~\ref{sec:results}, we present results only for discordant, dichotomous environmental effects, as the other results are very similar.

@@

We begin by comparing the ordinary lasso, PC-lasso, and LMM-lasso as the level of environmental confounding is varied. Estimation accuracy for lasso, PC-lasso, and LMM-lasso was evaluated based on mean squared error of the coefficients (MSE). Suppose $\bbetaHat(\lambda)$ is the vector of coefficients estimated by the lasso with a penalty of $\lambda$, and $\bbeta$ the coefficient values used to simulate the data. Mean squared error is then defined as
\begin{equation}
    \text{MSE} = \mathbb{E}||\bbeta - \bbetaHat(\lambda)||^2_2.
\end{equation}
Figure \ref{fig:mse} presents these results for all three data types in a 1:1 SNR setting, where $\eta = 0.5$. The top row of Figure \ref{fig:mse} shows that when $\xi = 0$ (i.e., no environmental effects), lasso and LMM-lasso perform comparably. PC-lasso results in a slight inflation in MSE compared to the other methods in the simulated data sets due to the fact that 10 additional predictors (PCs) have been added to the model, but none of them actually explain any variability in the outcome.

\begin{figure}[H]
    \centering
    \includegraphics[width = \textwidth]{figures/beta_mse.eps}
    \caption{Estimation accuracy of different methods for different data types in the presence and absence of environmental confounding with coarse subpopulation structure and dichotomous-discordant environmental effects.}
    \label{fig:mse}
\end{figure}

@@

In order to assess the behavior of these methods in this context, we conducted a simulation based on the Empirical genotype type, but restricted to just two subpopulations, Caucasian and African-American.  1000 SNPs were chosen from the genome-wide Empirical data such that the selected SNPs' absolute difference in MAF between the two subpopulations ranged uniformly from 0 to 0.5. As in the earlier simulations, the SNR was set to 1 ($\eta = 0.5$).

\begin{figure}[H]
\centering
\includegraphics[width = \textwidth]{figures/pop_inf_snps_box.eps}
\caption{Absolute allele frequency differences of selected, non-causal SNPs across methods, in the presence and absence of environmental confounding for two Empirical subpopulations, dichotomous-discordant effects, $\eta = 0.5$.}
\label{fig:pop_inf}
\end{figure}

@@

@@

Although PC-lasso and LMM-lasso outperform the ordinary lasso in the presence of environmental confounding, they are not equally effective across all population structures.  Here, we compare the performance of these two methods in two settings, as described in Section~\ref{sec:subpopulation_structure}: ``coarse'' (4 subpopulations) and ``fine'' (100 subpopulations).  Synthetic phenotypes were simulated using a 1:1 SNR ($\eta = 0.5$) and varying levels of environmental confounding, $\xi \in \{0.2, 0.5,0.8\}$.

\begin{figure}[H]
    \centering
    \includegraphics[scale = 0.9]{figures/mse_diff_subpops.eps}
    \caption{Relative performance of LMM-lasso, PC-lasso for fine and coarse subpopulation structures and varying levels of environmental confounding with Independent Subpopulations data, $\eta = 0.5$, and dichotomous-discordant environmental effects. }
    \label{fig:big_vs_small}
\end{figure}

@@

The underlying reason for this is that the RRM becomes ill-conditioned when $n > p$ \citep{ledoit2004well}. Although both approaches utilize the RRM, because LMM-lasso-ggmix attempts to estimate more parameters than LMM-lasso-Rakitsch, its optimization problem is no longer convex, making it prone to becoming stuck in local minima before it is able to sufficiently explore the coefficient path. This phenomenon is illustrated in greater detail in Supplementary Figure \ref{fig:niter}.

\begin{figure}[H]
    \centering
    \includegraphics[width = \textwidth]{figures/eta_beta_hat.eps}
     \caption{$\bbeta$ and $\eta$ estimation accuracy of LMM-lasso-Rakitsch and LMM-lasso-ggmix for varying $\eta$ levels and sample sizes with Independent Subpopulations data, coarse subpopulation structure, $\xi = 0.8$, and dichotomous-discordant environmental effect structure.}
    \label{fig:eta_beta_mse}
\end{figure}
